 

### 2021-01-11 vertical-thermal-review.Rmd
Update vertical-thermal-review.Rmd
Comments and edits up to line 430 / Ecosystem Scaling & Implications Section

Tropical forests with densely packed overstory trees limit light tosubsequent canopy layers and forest understories. Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]. However, vertical canopy layers and forest floors also receive light permeating through canopy gaps and intermittent light penetration from crown, branch, and leaf movement caused by wind called sunflecks. Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory depending on canopy attributes such as foliage density, height and architecture [Way and Pearcy 2012, REF]

@@

**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows of a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a] and in the northwestern United States [@davis_microclimatic_2019]. Even though canopy cover buffers maximum air temperatures in the understory, inevitable increase in understory temperatures continue to be observed as a result of global increase in temperatures, however the buffering by canopy cover smoothens this warming in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020;@rollinsonClimateSensitivityUnderstory]

Diurnal moderating capacity differs between forests with varying canopy structure, crown density, and seasons. Tropical, temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures than pine and boreal forests. Increase in crown density and leaf area index decreases light and higher maximum temperature permeability into forest understory, thus maintaining greater lower maximum daytime temperature. Comparitively open forests such as pine forests have sparse stands resulting in decreased buffering capacity in understory and closer air temperatures to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a]. This relationship is reveresed during the night, however, with pine forests maintaining warmer below canopy temperature than above[@vonarxSpatiotemporalEffectsForest2012a]

Seasonally, maximum air temperatures were higher above than below tropical forest canopies during wet seasons in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal Brazil [@fauset_differences_2018, Tymen et al. 2017].Similarly, maximum temps were higher and humidity lower in gaps vs understory regions (both at 1.5 m above the ground, see Fig. 4 - again illustrating buffering effect of the canopy).However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more air movement between subcanopy and above-canopy air[@rey-sanchez_spatial_2016], similar results were observed in in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].

Minimum $T_{air}$ is also buffered by forest canopies under some conditions.
Higher minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].(Martijn: should we refer to mechanisms at this point? (e.g., greater radiative heat loss in exposed areas resulting in lower Tmin))
However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].

@@

**Humidity also varies across the forest vertical profile, being generally higher in the understory (Fig. `r fig_NEON_vertical`).**
- DETAILS - REFS
- @mcgregor_tree_nodate

Along with temperature, forest canopies moderate relative humidity in below-canopy microclimates. In three European forest types (Broadleaved, Non-Pine Conifer, Pine) canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area. Moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].

RH, in combination with temperature, determines vapor pressure deficit, $VPD$. Atlantic forests during wet season tended to have higher $T_{air}$, and $VPD$ in upper canopy and canopy gaps than lower in understory where $T_{air}$ and $VPD$ were lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].  
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential. [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]. In North Eastern US temperate forests, upper canopy was observed to be more sensitive to $VPD$, whereas understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]

**Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
	Higher in understory, particularly at dusk [@koike_leaf_2001]. Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999]. Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]

Given that differences are small during the day when photosynthesis is active, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentration is unlikely to have much effect on the energy balance and metabolism of leaves across the forest vertical gradient.

If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO2 emitted at the forest floor could increase the vertical CO2 gradient in the future, but this would still only be a short-term effect in the early morning, as the CO2 would quickly diffuse through the canopy. Furthermore, increases in mortality of big trees (e.g. Senf et al. 2018) would increase canopy roughness and prevent build-up of CO2 concentration gradients.

@@

**The strength of this buffering varies across forests, being influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.**
First, buffering increases with canopy cover and water availabity.
* greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
* greater cover -->  lower max T and light availability [@zellweger_seasonal_2019]
* [@thom_effects_2020; Tymen et al. 2017 ; Jucker et al. 2018 ; Hardwick et al. 2015] 
The omega factor (McNaughton and Jarvis, 1983; Jarvis & McNaughton 1986) describes the degree of aerodynamic coupling between the vegetation and the atmospheric boundary layer; the lower the value of omega, the greater the coupling.
"I wonder whether it would be useful to introduce such 'formal' terms used to describe what is essentially the same as buffering effect of canopies.
I'd have to read up some more to see if there are papers showing what sort of canopy characteristics influence omega.

@@

Leaf traits are shaped on the basis of light and height and show different degrees of foliar plasticity in accordance with the intensity and availability of light which also optimizes water uptake, heat and C uptake by leaves across forest vertical strata. After the formation of leaves to integrated light, it takes 30-60 days for traits to stabalize in woody and herbaceous species [@niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves such as phytochromes among others mediate response to PAR, and ratio of red to far red radiation [R:FR] falling on the surface. This enables acclimation or avoidance response through development of necessary traits [Casal, 2013]. As a result, degrees of high irradiance gives rise to variations of sun-leaf traits  compared to degrees of low irradiance-- characterized by shade-- shapes variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also shapes leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al. 2016].

Increasing number of studies point towards the importance for incorporating sun and shade in modelling leaf economic spectrum (LES) within-canopy gradient[@chenLeafEconomicsSpectrum2020a] because LES was 'built' using mainly sun (upper canopy) leaves; given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016]. However, there is also concern for ambiguity in simplified cateogorization of 'sun' and 'shade' leaves due to variation based on light within-canopy, across functional groups, in canopy gaps, height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018]. Classification for sun and shade tolerance rankings have been highlighted by multiple studies [Baker 1949; @niinemetsControversyTraitsConferring2006; Feng et al. 2018] such as from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%". "Additionally, shade leaves in the understory are of three categories; 1. shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded. 2. seedlings/saplings of canopy trees that will eventually, if they survive, reach the canopy. 3 understory plants that complete their lifecycle in the shaded understory. These groups are likely to have different traits, and if we expect'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered".[refs, -Martijn Slot]

@@

Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees shaded by overstory. As a result, sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and more lobated than shade leaves
[Rozendaal et al. 2006, Mathur et al. 2018, more REFS].

However, the combination of stressors taller trees' foliage experience are different from the combination of stressors understory shorter trees's foliage experience. For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, wind speed, etc. whereas for understory--shade or limitation of light, niche competition, herbivory etc. These pressures also contribute to the development of necessary traits that is primarily mediated by light response. Overstory sun leaves are subjected to overheating pressures, thus through developed traits of increased stomatal density, transpiration and water supply along with increased convective heatloss by reducing boundary layer resistance, sun leaves are able to maintain optimum temperature. Additionally, overstory sun leaves tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Adaptive traits in understory enable light capture efficiency while maintaing larger width leaves with comparitevely fewer stomtal density as a result of available water, lesser $VPD$, and wind speed. "Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn].Those differences in traits are likely to affect their thermal properties as well.Leaves act as windows to plant performance enhacing plant adaptation to changing environment, thus traits acclimated to prior stressors have been suggested to enable whole plant to respond differently to forcoming stress [Rozendaal et al. 2006; @niinemetsResponsesForestTrees2010a]

However, even with distinct features of sun and shade traits, trait adaptation to light is done in such a way where light use efficiency for photosynthesis is maximized at both high or low light, while also adapting to other stressors as mentioned above. In light limited conditions, shade leaves increase light capture efficiency through increasing their specific leaf area [SLA] and by investing in chlorophyll at a mass basis--"leads to a similar chlorphyll content on an area basis compared with sun leaves. This makes shade leaves realize a similar light capture to sun leaves at a lower biomass investment" [Chazdon et al. 1996, Poorter et al. 2000, Rozendaal et al. 2006]

With increasing height and age, competion for light decreases as availibity  of light increases in dense forests. However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements. Early successional species are considered more shade intolerant than late successional species that are more shade tolerant. Ontogenic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$) in shade tolerators than in intolerators, greater photosynthetic capacity in low light and greater carbon gain capacities[Niinemets, 2006].Rozendaal et al. 2006, Koike et al. 2001 has more info about this.

@@

**Many traits vary with height and/or between sun and shade leaves on the same species (Table 1).**
Here, we focus on intra-canopy traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.

@@

Are traits shaped more by height or light?
Height is more important than light in determining leaf morphology in a tropical forests (Cavaleri et al. 2010)
- (Cavaleri et al. 2008) 
- from Martijn: "If I remember correctly, height and light exposure are closely correlated and things like LMA scale with both, but above a certain level of light exposure, the effect of additional light disappears, while the relationship of the morphological traits with height continues. I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
-"Our study suggests that height plays an important role in the leaf traits, in contrast with previous findings (Baranski, 1975; Sack et al., 2006).In Quercus. alba and Quercus. velutina, loba-tion decreased with increasing tree height. Lobation allows air passage and light penetration in the tree canopy (Sack et al., 2006; Semchenko & Zobel, 2007)" [@kusiPlasticLeafMorphology2020]

@@

*Isoprene production*
-  Isoprene production *capability* as a trait (actual isoprene production is in section 4. I'm not sure whether Ty's content belongs here or in that section)

@@

**Many of the biophysical and trait variable reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019] (Fig. 3).
- As light availability typically increases with tree height, leaf temperature which is largely of function of radiation input, does as well [@michaletzPlantThermoregulationEnergetics2015]. Direct sunlight in upper canopy leaves increases leaf temperature, [sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress] more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature.[@fauset_differences_2018]
Large leaves can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation with stomata closed.  
Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm).
Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
However, counteracting this, lower wind speed in understory would reduce latent heat loss.
Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance; however, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
- increase in air temperature, increases VPD, decrease in stomtal conductance during miday leading to decrease in CO2 in leaf = increase in leaf temperature in upper canopy. [@fauset_differences_2018]

@@

**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. **2?**).**
In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999).
Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil Forests despite much lower air temperature due to high solar radiation loads. However, leaf temperatures themselves were moderaed by air temperature. [@fauset_differences_2018].
-	(Slot et al. 2019) and refs therein
-	(Fauset et al. 2018)
Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.

*(Because hydraulics are key to leaf T, we should include some of that here--e.g., water on leaves and leaf water content)*

>>>>

Tropical forests with densely packed overstory trees limit light tosubsequent canopy layers and forest understories. Tropical forest floors receive approximately 1-2%  of the incident light that hits the top of the canopy [@robertsStomatalBoundaryLayerConductances1990a; @chazdonPhotosyntheticLightEnvironments1984; @parkerConsequencesEnvironmentalHeterogeneity2019], whereas mixed-wood boreal and Northern temperate forest floors receive around 4-5% of full sunlight [@bartemucciOverstoryInfluencesLight2006]. However, vertical canopy layers and forest floors also receive light permeating through canopy gaps and intermittent light penetration from tree, branch, and leaf movement caused by wind called sunflecks. Sunflecks can provide 10-80% of photon flux density for photosynthesis in canopy layers and understory depending on canopy attributes such as foliage density, height and architecture [Way and Pearcy 2012, REF]

@@

**Air temperature, $T_{air}$, is sometimes significantly buffered by forest canopies (Fig. `r fig_NEON_vertical`).**
Studies comparing $T_{air}$ under forest canopies with nearby clearings have found lower maximum temperatures under forest canopies globally [@defrenneGlobalBufferingTemperatures2019], and recent evidence shows a similar trend across Europe [@zellweger_seasonal_2019; @vonarxSpatiotemporalEffectsForest2012a] and in the northwestern United States [@davis_microclimatic_2019]. Intact tropical forests have also been show to exhibit lower peak daytime $T_{air}$ than neighboring tree crop agricultural plantations (S. R. Hardwick et al., The relationship between leaf area index and microclimate in tropical forest and oil palm plantation: Forest disturbance drives changes in microclimate. Agric. Meteorol. 201, 187–195 (2015)). Even though canopy cover buffers maximum air temperatures in the understory, inevitable increases in understory temperatures continue to be observed as a result of global temperature increases. However, the buffering by canopy cover smooths this warming in the understory [@bertrandCommentForestMicroclimate2020; @zellwegerResponseCommentForest2020;@rollinsonClimateSensitivityUnderstory]

Diurnal moderating capacity differs between forests with varying canopy structure, crown density, and seasonality. Tropical, temperate broadleaved and non-pine conifer forests maintain lower daytime maximum temperatures than pine and boreal forests. Increases in crown density and leaf area index decrease light and higher maximum temperature permeability into the forest understory, thus maintaining lower maximum daytime temperatures. Comparitively open forests such as pine forests with sparse stands (stems?) resulting in decreased buffering capacity in the understory and air temperatures more similar to nearby clearings [@defrenneGlobalBufferingTemperatures2019; @vonarxSpatiotemporalEffectsForest2012a]. This relationship is reveresed during the night, however, with open-canopy pine forests maintaining warmer understory temperatures than above canopy [@vonarxSpatiotemporalEffectsForest2012a]

During the wet season, maximum air temperatures have been shown to be higher above forest canopies than below in tropical forests in Panama [@rey-sanchez_spatial_2016], temperate forests in Switzerland [@vonarxSpatiotemporalEffectsForest2012a] and coastal forests in Brazil [@fauset_differences_2018, Tymen et al. 2017]. Similarly, in these studies, maximum temperatures were higher while humidity was lower in gaps compared to understory regions (both at 1.5 m above the ground, see Fig. 4 - again illustrating buffering effect of the canopy). However, similar maximum temperatures have been observed during the dry season of a semi-deciduous tropical forest in Panama, partially because in the dry season some canopy trees were leafless, resulting in more light reaching the sub-canopy, and more (Elsa: replace end of sentence with 'boundary layer turbulence'?) air movement between subcanopy and above-canopy air [@rey-sanchez_spatial_2016]. Similar results were observed in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate].

Minimum $T_{air}$ is also buffered by forest canopies under some conditions.
Higher (Elsa: or lower?) minimum temperatures under forest canopies relative to nearby clearings have been observed across Europe [@zellweger_seasonal_2019] and in the northwestern United States [@davis_microclimatic_2019].(Martijn: should we refer to mechanisms at this point? (e.g., greater radiative heat loss in exposed areas resulting in lower Tmin))
However, similar minimum temperatures under forest canopies have been observed in tropical forests in Panama [@rey-sanchez_spatial_2016] and coastal Brazil [@fauset_differences_2018], as well as in a temperate deciduous forest in the eastern United States [@mcgregor_tree_nodate; @nakamuraForestsTheirCanopies2017].

@@

**Humidity also varies across the forest vertical profile, and is generally higher in the understory (Fig. `r fig_NEON_vertical`).**
- DETAILS - REFS
- @mcgregor_tree_nodate

Along with temperature, forest canopies moderate relative humidity (RH) in below-canopy microclimates. In three European forest types (Broadleaved, Non-Pine Conifer, Pine), canopy buffering decreased daily maximum $T_{air}$ and increased daily minimum RH in below-canopy microclimate compared to a nearby open area. Moderating capacity was stronger under broadleaved and non-pine conifer forests than pine, and greater in wetter conditions, and in summer and autumn months than winter [@vonarxSpatiotemporalEffectsForest2012a].

RH, in combination with temperature, determines vapor pressure deficit, $VPD$. Atlantic forests during the wet season tend to have higher $T_{air}$, and $VPD$ in the upper canopy and canopy gaps than lower in the understory where $T_{air}$ and $VPD$ tend to be lower [@fauset_differences_2018; @tymenQuantifyingMicroenvironmentalVariation2017].  
This means that canopy leaves tend to be exposed to higher *evaporative demand*, higher stomtal contrains and lower osmotic potential. [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]. In North Eastern US temperate forests, the upper canopy was observed to be more sensitive to $VPD$, whereas the understory was insensitive [@rollinsonClimateSensitivityUnderstory, 2020]

**Finally, carbon dioxide (CO~2~) concentrations tend to be higher in the understory.** 
	Higher in the understory, particularly at dusk [@koike_leaf_2001]. Higher in the understory overnight; difference persists during the day but is very small [@yang_spatial_1999]. Differences in concentration are by far most pronounced near ground level, as well, concentrations of (CO~2~) mole fraction values have been observed to be higher in lower canopy leaves compared with upper canopy leaves [@niinemetsPhotosyntheticAcclimationSimultaneous2004a]

Given that differences are small during the day when photosynthesis is occuring, and that even nighttime differences are modest at the height of understory tree crowns, CO~2~ concentrations are unlikely to have much of an effect on the energy balance and metabolism of leaves across the forest vertical gradient.

If soil and stem respiration do not acclimate fully to rising temperatures, the greater amount of CO~2~ emitted at the forest floor (i.e. from soils and decaying organic matter and leaf litter) could increase the vertical CO~2~ gradient in the future, but this would still only be a short-term effect in the early morning, as the CO~2~ would quickly diffuse through the canopy. Furthermore, increases in mortality of big trees (e.g. Senf et al. 2018) would increase canopy roughness and prevent build-up of CO~2~ concentration gradients.
(Elsa: see these papers on the long-term Harvard Forest soil warming experiments. CO2 emissions increased, then decreased, then increased again. It's thought that the changes in soil CO2 emissions over time were linked to turnover in microbial community composition.
- DeAngelis, K.M., Pold, G., Topçuoğlu, B.D., van Diepen, L.T., Varney, R.M., Blanchard, J.L., Melillo, J. and Frey, S.D., 2015. Long-term forest soil warming alters microbial communities in temperate forest soils. Frontiers in microbiology, 6, p.104.
- Melillo, J.M., Steudler, P.A., Aber, J.D., Newkirk, K., Lux, H., Bowles, F.P., Catricala, C., Magill, A., Ahrens, T. and Morrisseau, S., 2002. Soil warming and carbon-cycle feedbacks to the climate system. Science, 298(5601), pp.2173-2176.
- Melillo, J.M., Frey, S.D., DeAngelis, K.M., Werner, W.J., Bernard, M.J., Bowles, F.P., Pold, G., Knorr, M.A. and Grandy, A.S., 2017. Long-term pattern and magnitude of soil carbon feedback to the climate system in a warming world. Science, 358(6359), pp.101-105.)

@@

**The strength of this buffering varies across forests, influenced by both forest characteristics (Fig. `r fig_NEON_differences`) and the biophysical environment.**
First, buffering increases with canopy cover and water availabity.
* greater cover --> lower max T and VPD, higher minT  [@davis_microclimatic_2019]
* greater cover -->  lower max T and light availability [@zellweger_seasonal_2019]
* [@thom_effects_2020; Tymen et al. 2017 ; Jucker et al. 2018 ; Hardwick et al. 2015] 
The omega factor (McNaughton and Jarvis, 1983; Jarvis & McNaughton 1986) describes the degree of aerodynamic coupling between the vegetation and the atmospheric boundary layer; the lower the value of omega, the greater the coupling.
"I wonder whether it would be useful to introduce such 'formal' terms used to describe what is essentially the same as buffering effect of canopies.
I'd have to read up some more to see if there are papers showing what sort of canopy characteristics influence omega.

@@

Leaf traits are shaped on the basis of light and height and show different degrees of intraspecific foliar plasticity in accordance with the intensity and availability of light which also optimizes water uptake, heat and C uptake by leaves across forest vertical strata. After the formation of leaves to integrated light, it takes 30-60 days for traits to stabalize in woody and herbaceous species [@niinemetsLeafAgeDependent2016a]. A network of sensory photoreceptors in leaves such as phytochromes among others mediate leaf-level responses to PAR, and ratio of red to far red radiation [R:FR] intercepting the leaf surface. This enables acclimation or avoidance responses through the development of necessary leaf traits [Casal, 2013]. As a result, degrees of high irradiance gives rise to variation in sun-leaf traits compared to degrees of low irradiance-- characterized by shade-- shapes variations of shade-leaf traits [@poorterMetaanalysisPlantResponses2019b; Sack et al 2006], which ultimately also shapes (Elsa: I'd recommend using influence or some other word instead of 'shapes' given that leaf shape is a trait sometimes measured and so it may be confusing in this context) leaf structural, chemical and physiological traits as well [Keenan and Niinemets et al. 2016].

(Elsa: it may also be worth mentioning leaf turnover / leaf longevity briefly in the above paragraph in the context of leaf development and trait formation.)

An increasing number of studies point towards the importance of incorporating sun and shade in modelling leaf economic spectrum (LES) (traits?) within-canopy gradient [@chenLeafEconomicsSpectrum2020a] because LES theory was developed using mainly sun exposed (upper canopy) leaves. Given the vastly different light environments at the bottom of the canopy (and that most leaves in the canopy are likely to be shade leaves) there is a strong research need to better characterise trait relationships and leaf responses through the whole vertical canopy [Keenan and Niinemets 2016]. However, there is also concern for ambiguity in a simplified cateogorization of 'sun' and 'shade' leaves due to variation based on within-canopy light, across functional groups, in canopy gaps, wtih height and ontogeny [@ishiiVariationIntracrownLeaf2018; @niinemetsWorldwideAnalysisWithincanopy2015] as well as geographically [@ishiiVariationIntracrownLeaf2018]. Classification schemes for sun and shade tolerance rankings have been highlighted by multiple studies [Baker 1949; @niinemetsControversyTraitsConferring2006; Feng et al. 2018]. For example, classifications from @niinemetsControversyTraitsConferring2006 are as follows: "The five-level scale used for shade tolerance (1,very intolerant; 2, intolerant; 3, moderately tolerant; 4,tolerant; 5, very tolerant) corresponds approximately to the following light availabilities expressed as percentage of full sunlight: 1,>50%; 2, 25-50%; 3, 10-25%; 4, 5-10%; 5, 2-5%". "Additionally, shade leaves in the understory are of three categories; 1. shade leaves of canopy trees--top of the tree is sun exposed, bottom is shaded. 2. seedlings/saplings of canopy trees that will eventually, if they survive, reach the canopy. 3 understory plants that complete their lifecycle in the shaded understory. These groups are likely to have different traits, and if we expect 'shade' plants to have a different sensitivity to climate change than sun leaves, the consequences of climate change for demography and species community composition of the forest will depend strongly on the category of shade plants being considered".[refs, -Martijn Slot]

@@

Taller trees in dense forests tend to be exposed to greater irradiance than lower canopy and shorter understory trees that are shaded by overstory. As a result, the sun and shade leaf trait dichotomy has been established across species by various studies where typical sun leaves are often smaller in size and width, thicker, hairier, with thicker cuticle and more lobated than shade leaves
[Rozendaal et al. 2006, Mathur et al. 2018, more REFS].

However, the combination of stressors taller trees' foliage experience are different from the combination of stressors understory trees' foliage experience. For example, overstory sun leaves experience greater loads of PAR, $VPD$, elevated temperatures, higher wind speed, etc. whereas for understory--shade or limitation of light, niche competition, herbivory etc. These pressures also contribute to the development of necessary traits that are primarily mediated by light response. Overstory sun leaves are subjected to overheating pressures, thus through developed traits of increased stomatal density, transpiration and water supply along with increased convective heatloss by reducing boundary layer resistance, sun leaves are able to maintain optimum temperature. Additionally, overstory sun leaves tend to have long-lived leaves with high LMA while shade leaves of canopy trees have large leaves with low LMA. Adaptive traits in understory enable light capture efficiency while maintaing larger width leaves with comparitevely fewer stomtal density as a result of available water, lesser $VPD$, and wind speed. "Spending their lifecycle in the shade means they have different pressures than plants that get the bulk of their carbon from sun-exposed leaves, but can maintain a few layers of shade leaves as well."[-Martijn]. Those differences in traits are likely to affect their thermal properties as well. Leaves act as windows to plant performance, enhacing plant adaptation to changing environment, thus traits acclimated to prior stressors have been suggested to enable whole plants to respond differently to forcoming stress [Rozendaal et al. 2006; @niinemetsResponsesForestTrees2010a]

However, even with distinct features of sun and shade traits, trait adaptation to light occurs in such a way where light use efficiency for photosynthesis is maximized at both high or low light, while also adapting to other stressors as mentioned above. In light limited conditions, shade leaves increase light capture efficiency through increasing their specific leaf area [SLA] and by investing in chlorophyll on a mass basis--"leads to a similar chlorphyll content on an area basis compared with sun leaves. This enables shade leaves to realize similar light capture to sun leaves at a lower biomass investment" [Chazdon et al. 1996, Poorter et al. 2000, Rozendaal et al. 2006]

With increasing height and age, competition for light decreases as availibity  of light increases in dense forests. However, depending on growth strategies of species occupying the understory, leaf traits might differ due to species inherent light requirements. Early successional species are considered more shade intolerant than late successional species that are more shade tolerant. Ontogenetic differences between the two in the understory suggests lower leaf dry mass per area ($M_A$) in shade tolerators than in intolerators, greater photosynthetic capacity in low light and greater carbon gain capacities[Niinemets, 2006].Rozendaal et al. 2006, Koike et al. 2001 has more info about this.

@@

**Many traits vary with height and/or between sun and shade leaves of the same species (Table 1).**
Here, we focus on intra-canopy traits that are important for shaping leaf temperature ($T_{leaf}$) and thermal sensitivity.

@@

Are traits shaped more by height or light?
Height is more important than light in determining leaf morphology in a tropical forests (Cavaleri et al. 2010)
- (Cavaleri et al. 2008) 
- from Martijn: "If I remember correctly, height and light exposure are closely correlated and things like LMA scale with both, but above a certain level of light exposure, the effect of additional light disappears, while the relationship of the morphological traits with height continues. I have often been a bit skeptical about assuming height drives morphology, as most of the work that convincingly shows that (and relates it to turgor pressure for example) is from 100 m tall redwoods."
-"Our study suggests that height plays an important role in the leaf traits, in contrast with previous findings (Baranski, 1975; Sack et al., 2006).In Quercus. alba and Quercus. velutina, loba-tion decreased with increasing tree height. Lobation allows air passage and light penetration in the tree canopy (Sack et al., 2006; Semchenko & Zobel, 2007)" [@kusiPlasticLeafMorphology2020]
- Elsa: see also H. Poorter, U. Niinemets, L. Poorter, I. J. Wright, R. Villar, Causes and consequences of variation in leaf mass per area (LMA): A meta-analysis. New Phytol. 182, 565–588 (2009). -- LMA responds to changes in light, with shade-intolerant species displaying higher plasticity in response to light.

@@

*Isoprene production*
-  Isoprene production *capability* as a trait (actual isoprene production is in section 4. I'm not sure whether Ty's content belongs here or in that section) 
- Elsa: I think Ty's section on isoprene would be better suited up here

@@

**Many of the biophysical and trait variable reviewed above affect leaf temperature, $T_{leaf}$, which in turn has a strong influence on leaf metabolism (see following section).**
$T_{leaf}$ is determined by the energy balance of a leaf and can be estimated based on basic biophysical principles [@campbell_introduction_1998; @muir_tealeaves_2019] (Fig. 3).
- As light availability typically increases with tree height, leaf temperature which is largely a function of radiation input, does as well [@michaletzPlantThermoregulationEnergetics2015]. Direct sunlight in upper canopy leaves increases leaf temperature, [sometimes 1-7C above ambient air temperature, making upper canopy leaves susceptible to heat stress] more than scattered light in lower shaded leaves [@mauTemperateTropicalForest2018a]
While small leaves remain within a few degrees of $T_{air}$, regardless of stomatal conductance, wider leaves can have temperatures deviating more from air temperature.[@fauset_differences_2018]
Large leaves can be significantly cooler than $T_{air}$ under low radiation with stomata open, and significantly hotter under high radiation with stomata closed.  
Leaves with open stomata and high radiation loads maintain $T_{leaf}$ similar to $T_{air}$, with coolest leaves at intermediate sizes (~10mm).
Shaded understory leaves should tend to maintain cooler daytime leaf temperatures for any given level of stomatal conductance.
However, counteracting this, lower wind speed in understory would reduce latent heat loss.
Thus, under hot conditions, canopy leaves exposed to higher wind speeds would be most effective at cooling when sufficient water is available to maintain high stomatal conductance; however, their $T_{leaf}$ can be highly elevated above $T_{air}$ when stomatal conductance is limiting.
- increase in air temperature, increases VPD, decrease in stomatal conductance during midday leading to decrease in CO2 in leaf = increase in leaf temperature in upper canopy. [@fauset_differences_2018]

@@

**When vertical gradients in the biophysical environment and leaf traits combine to shape leaf energy budgets, under some conditions $T_{leaf}$ tracks  $T_{air}$ and therefore differs little between understory and canopy (Fig. **2?**).**
In a broadleaf deciduous forest, nighttime $T_{leaf}$'s were similar in the canopy and understory	(Bolstad et al. 1999).
Similarly, during the dry season in a tropical moist forest in Panama... (Rey-Sánchez et al. 2016).
**However, canopy leaves can reach much higher maximum $T_{leaf}$, and higher $T_{leaf} - T_{air}$, than understory leaves.**
-Large leaf to air temperature differences and high leaf temperatures were found in upper canopy of Atlantic, Brazil Forests despite much lower air temperature due to high solar radiation loads. However, leaf temperatures themselves were moderated by air temperature. [@fauset_differences_2018].
-	(Slot et al. 2019) and refs therein
-	(Fauset et al. 2018)
Sun leaves can also be  cooled relative to $T_{air}$ more than shade leaves (Rey-Sánchez et al. 2016). There's some hint of this in Bolstad et al. 1999.

*(Because hydraulics are key to leaf T, we should include some of that here--e.g., water on leaves - evaporative cooling - and leaf water content)*
